---
title: "P8105 Midterm"
author: "Mengfan Luo (ml4701)"
date: "10/30/2021"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


### 1 - Data

```{r}
# Load and clean variable names
eop_df = read_excel("p8105_mtp_data.xlsx",range = "A9:I1230") %>% 
  janitor::clean_names() %>% 
  rename(eop_visibility = eop_visibility_classification) %>% 
  mutate(
    eop_size_mm = str_replace_na(eop_size_mm,"0"), # Filled missing values in eop_size
    
    #data type conversion
    sex = factor(sex),
    sex = recode(sex,"0" = "female", "1" = "male"),
    
    age_group = fct_inseq(age_group,ordered = TRUE),
    eop_size_mm = as.numeric(eop_size_mm),
    eop_size = fct_inseq(eop_size,ordered = TRUE),
    eop_visibility = fct_inseq(factor(eop_visibility),ordered = TRUE),
    eop_shape = factor(eop_shape),
    fhp_category = fct_inseq(fhp_category,ordered = TRUE))

```


The following steps are done to clean and tidy up the data:

1. Columns names are converted to lower snake case

2. Missing values in `eop_size_mm` are filled with `0` as indicated in the original dataset (Yet missing values in `eop_shape` and `fhp_size` are kept as `NA` as no such information provided).

3. `eop_size_mm` is converted to numeric variable; `sex`,`eop_shape`,`age_group`, `eop_size`,`eop_visibility` and `fhp_category` are converted to factors.

The resulting dataset has `r nrow(eop_df)` rows and `r ncol(eop_df)` columns. Based on the original scientific report,

**key variables:**

* `sex`: gender of participants, factor with 2 unordered levels `r levels(pull(eop_df,sex))`
* `age`: age of participants, numeric variable with mean `r mean(pull(eop_df,age))` and range [`r range(pull(eop_df,age))`]
* `age_group`: age groups divided according to decades. Has 8 ascending levels `r levels(pull(eop_df,age_group))`
* `eop_size_mm`: size of EOP, numeric variable with mean `r mean(pull(eop_df,eop_size_mm))` and range [`r range(pull(eop_df,eop_size_mm))`]. 
* `eop_size`: ordered factor encoded from `eop_size_mm` with 7 ascending levels `r levels(pull(eop_df,eop_size))`, where "14.6" seems to be a typo in the dataset. 
*Note: `eop_size` > 1 (`eop_size_mm`> 10) were defined as enlarged EOP (EEOP).*
* `fhp_size_mm`: size of FHP, numeric variable with mean `r mean(pull(eop_df,fhp_size_mm),na.rm = TRUE)` and range [`r range(pull(eop_df,fhp_size_mm),na.rm = TRUE)`], has `r sum(is.na(pull(eop_df,fhp_size_mm)))` missing values. 
* `fhp_category`: ordered factor encoded from `fhp_size_mm` with 9 ascending levels `r levels(pull(eop_df,fhp_category))`. 

**Other variables:**

* `eop_visibility`: factor with 3 ordered levels `r levels(pull(eop_df,eop_visibility))`
* `eop_shape`:  factor with 4 unordered levels `r levels(pull(eop_df,eop_shape))`, has `r sum(is.na(pull(eop_df,eop_shape)))` missing values

Despite `r sum(is.na(eop_df))` missing values, most of them (`r sum(is.na(eop_df$eop_shape))`) lie in `eop_shape` and is not key variables in the paper, so a total of `r nrow(eop_df)` participants are included.

The age and gender distribution are displayed with following table and plot:

```{r}
eop_df %>% 
  mutate(age_group = recode(age_group,"1" = "<18","2" = "18-30", "3" = "31-40",
                            "4" = "41-50", "5" = "51-60", "6" = "61-70", 
                            "7" = "71-80", "8" = ">80")) %>% 
  janitor::tabyl(age_group,sex) %>% 
  knitr::kable()

eop_df %>%
  mutate(age_group = recode(age_group,"1" = "<18","2" = "18-30", "3" = "31-40",
                            "4" = "41-50", "5" = "51-60", "6" = "61-70", 
                            "7" = "71-80", "8" = ">80")) %>% 
  group_by(age_group,sex) %>% 
  ggplot(aes(x = age_group,fill = sex))+
  geom_bar(alpha = .7, position=position_dodge())+
  labs(title = "Gender and age distribution of Participants",
       x = "Age group",
       y = "Number of participants")
```


There's some internal inconsistency within the dataset. 

1. Definitions of categorical variables was not clearly specified, especially the endpoints of the intervals to partition continuous variables. For example, definition for `eop_size` based on `eop_size_mm` was given as:

| eop_size|eop_size_mm|
|--------:|:----------|
|0        | 0-5       |
|1        | 5-10      |
|2        | 10-15     |
|3        | 15-20     |
|4        | 20-25     |
|5        | 25+       |

But whether dividing points (5,10...) belong to the former or latter group was not specified, which results in problems shown in the following table, that both 15 and 20 are categorized into `eop_size` == 3 (and also a classifications into 14.6)


```{r}
eop_df %>% 
  select(eop_size_mm,eop_size) %>%
  filter(eop_size_mm %in% c(15,20)) %>% 
  knitr::kable()
```


2. Categorical variables in the dataset didn't correctly implement the definitions given. Some discordence between the *given categorical groups* and *recalculated groups* are shown in the following tables:

```{r}
eop_df %>% 
  select(age,age_group) %>%
  mutate(
         age_group_recalculate = case_when(
          age <= 17 ~ "1",
          (age > 17) & (age <= 30) ~ "2",
          (age > 30) & (age <= 40) ~ "3",
          (age > 40) & (age <= 50) ~ "4",
          (age > 50) & (age <= 60) ~ "5",
          (age > 60) & (age <= 70) ~ "6",
          (age > 70) & (age <= 80) ~ "7",
           age > 80 ~ "8")) %>% 
  filter(!(age_group == age_group_recalculate)) %>% 
  knitr::kable()
  

eop_df %>% 
  select(eop_size_mm,eop_size) %>%
  mutate(
         eop_size_recalculate = case_when(
      eop_size_mm <= 5 ~ "0",
      (eop_size_mm > 5 ) &(eop_size_mm <= 10) ~ "1",
      (eop_size_mm > 10) &(eop_size_mm <= 15) ~ "2",
      (eop_size_mm > 15) &(eop_size_mm <= 20) ~ "3",
      (eop_size_mm > 20) &(eop_size_mm <= 25) ~ "4",
      eop_size_mm > 25 ~ "5")) %>% 
  filter(!(eop_size == eop_size_recalculate))%>% 
  knitr::kable()

```


### 2 - Visualization

1. *Figure 3 (Improved)* displays distribution of FHP size across age and sex groups with violin plot,including major statistics quartiles and means. *Figure 4 (Improved)* shows the rate of EEOP across age and sex groups with a scatter plot.

```{r, message = FALSE}
fhp_mean = eop_df %>% 
  filter(age >= 18) %>% 
  mutate(
    age_group = case_when(
      age <= 30 ~ "2",
      (age > 30) &(age <= 40) ~ "3",
      (age > 40) &(age <= 50) ~ "4",
      (age > 50) &(age <= 60) ~ "5",
      (age > 60)~ "6"),
    age_group = factor(age_group, levels = c("2","3","4","5","6"),ordered = TRUE, 
                       labels = c("18-30 years","30s","40s","50s",">60"))) %>% 
  group_by(age_group,sex) %>% 
  summarize(fhp_mean = mean(fhp_size_mm,na.rm = TRUE))

fig3 = eop_df %>% 
  filter(age >= 18) %>% 
  mutate(
    age_group = case_when(
      age <= 30 ~ "2",
      (age > 30) &(age <= 40) ~ "3",
      (age > 40) &(age <= 50) ~ "4",
      (age > 50) &(age <= 60) ~ "5",
      (age > 60)~ "6"),
    age_group = factor(age_group, levels = c("2","3","4","5","6"),ordered = TRUE, 
                       labels = c("18-30 years","30s","40s","50s",">60"))) %>% 
  ggplot(aes(x = age_group,y = fhp_size_mm))+
  geom_violin(aes(fill = sex),na.rm = TRUE,
              draw_quantiles =c(0.25,0.5,0.75),
              width = 1, position = position_dodge(), 
              alpha = .3,size = .5)+
  geom_point(data = fhp_mean,aes(x = age_group, y = fhp_mean,shape = sex),
              size = 3,alpha = 1, position = position_dodge(width = 1))+
  labs(
    title= "Figure 3 (Improved). ",
    subtitle = "Forward head protraction values \nacross the age groups and sexes.",
       x = "Age Group (years)",
       y = "Forward Head Protraction Size (mm)")

fig4 = eop_df %>% 
  filter(age >= 18) %>% 
  mutate(
    age_group = case_when(
      age <= 30 ~ "2",
      (age > 30) &(age <= 40) ~ "3",
      (age > 40) &(age <= 50) ~ "4",
      (age > 50) &(age <= 60) ~ "5",
      (age > 60)~ "6"),
    age_group = factor(age_group, levels = c("2","3","4","5","6"),ordered = TRUE, 
                       labels = c("18-30 years","30s","40s","50s",">60")),
    eop_size = case_when(
      eop_size_mm <= 5 ~ 0,
      (eop_size_mm > 5 ) &(eop_size_mm <= 10) ~ 1,
      (eop_size_mm > 10) &(eop_size_mm <= 15) ~ 2,
      (eop_size_mm > 15) &(eop_size_mm <= 20) ~ 3,
      (eop_size_mm > 20) &(eop_size_mm <= 25) ~ 4,
      eop_size_mm > 25 ~ 5),
    eeop = case_when(eop_size <= 1 ~ 0,
                     eop_size > 1 ~1) 
    ) %>%
  group_by(age_group,sex) %>% 
  summarize(group_total = n(),EEOP_count = sum(eeop),
            rate = EEOP_count/group_total) %>% 
  ggplot(aes(x = age_group, y = rate,group = sex,shape = sex,color = sex))+
    geom_point(size = 3)+
    geom_line()+
    labs(
      title= "Figure 4 (Improved). ",
      subtitle = "Rate of the presence of EEOP \nacross the age groups and sexes.",
         x = "Age Group (years)",y = "EEOP Count")

fig3+fig4

```

2. From the following plot, no clear trend is seen either accross `age` and `sex` groups, or between `eop_size_mm` and `fhp_size_mm`, so EOP size may not be associated with FHP size, gender or age.

```{r}
eop_df %>% 
  select(age,sex,eop_size_mm,fhp_size_mm) %>% 
  drop_na() %>%
  filter(age >= 18) %>% 
  mutate(
    age_group = case_when(
      age <= 30 ~ "2",
      (age > 30) &(age <= 40) ~ "3",
      (age > 40) &(age <= 50) ~ "4",
      (age > 50) &(age <= 60) ~ "5",
      (age > 60)~ "6"),
    age_group = factor(age_group, levels = c("2","3","4","5","6"),ordered = TRUE, 
                       labels = c("18-30 years","30s","40s","50s",">60")),
    eop_size = case_when(
      eop_size_mm <= 5 ~ 0,
      (eop_size_mm > 5 ) &(eop_size_mm <= 10) ~ 1,
      (eop_size_mm > 10) &(eop_size_mm <= 15) ~ 2,
      (eop_size_mm > 15) &(eop_size_mm <= 20) ~ 3,
      (eop_size_mm > 20) &(eop_size_mm <= 25) ~ 4,
      eop_size_mm > 25 ~ 5),
    eeop = case_when(eop_size <= 1 ~ "non-EEOP",
                     eop_size > 1 ~"EEOP") ) %>%
  ggplot(aes(x = fhp_size_mm,y = eop_size_mm,color = eeop))+
    geom_point(alpha = .6,size = 2)+
    facet_grid(sex~age_group)+
    labs(
      subtitle= "Association between FHP size and EOP size across age and sex groups ",
         x = "FHP size (mm)",y = "EOP size (mm)",color = "EEOP status")


```






```{r}
wordcountaddin::text_stats("p8105_mtp_ml4701.Rmd")
```

































